---
title: "accessibletablesR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{accessibletablesR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is a demonstration of how to use accessibletablesR.  
<br>
accessibletablesR is designed to produce Excel workbooks that align as closely as possible to the UK Government Analysis Function recommendations for publishing statistics in spreadsheets.  
<br>
[Releasing statistics in spreadsheets](https://analysisfunction.civilservice.gov.uk/policy-store/releasing-statistics-in-spreadsheets/)  
<br>
accessibletablesR was developed using R 4.1.3, dplyr version 1.1.2, openxlsx version 4.2.5.2, conflicted version 1.2.0, devtools version 2.4.5, stringr version 1.5.0, purrr version 1.0.1 and odsconvertr version 0.2.2. It is unknown if the package will work with earlier versions of R, dplyr, openxlsx, conflicted, devtools, stringr, purrr and odsconvertr. accessibletablesR will install the latest versions of dplyr, openxlsx, conflicted, devtools, stringr, purrr and odsconvertr if these packages are not currently installed or if earlier versions of dplyr (<1.1.2), openxlsx (<4.2.5.2), conflicted (<1.2.0), devtools (<2.4.5), stringr (<1.5.0), purrr (<1.0.1) and odsconvertr (<0.2.2) 
are currently installed.

## Installation

To install accessibletablesR:

```{r, eval = FALSE}
if (!("devtools" %in% utils::installed.packages())) 
  {utils::install.packages("devtools", dependencies = TRUE, type = "binary")}

devtools::install_github("minifiemj/accessibletablesR", build_vignettes = TRUE)

# Optional - attach package
library("accessibletablesR")
```

If a firewall prevents install_github from working (a time out message may appear) then install the package manually. On the GitHub repo, go to the green "Code" icon and choose "Download ZIP". Copy 
the ZIP folder to a network drive. Use devtools::install_local(<link to the zipped folder>) to 
install the package.  
<br>
odsconvertr is also installed from GitHub. If a similar issue prevents the automatic installation of odsconvertr, then install manually from "department-for-transport/odsconvertr".

## Functions in the package

```{r, eval = FALSE}
workbook <- function(covertab = NULL, contentstab = NULL, notestab = NULL, autonotes = NULL,
                     definitionstab = NULL, fontnm = "Arial", fontcol = "black",
                     fontsz = 12, fontszst = 14, fontszt = 16, title = NULL, creator = NULL,
                     subject = NULL, category = NULL, autodel = NULL)
  
creatingtables <- function(title, subtitle = NULL, extraline1 = NULL, extraline2 = NULL, 
                           extraline3 = NULL, extraline4 = NULL, extraline5 = NULL, 
                           extraline6 = NULL, sheetname, table_data, headrowsize = NULL, 
                           numdatacols = NULL, numdatacolsdp = NULL, othdatacols = NULL, 
                           tablename = NULL, gridlines = "Yes", columnwidths = "R_auto", 
                           width_adj = NULL, colwid_spec = NULL)
  
contentstable <- function(gridlines = "Yes", colwid_spec = NULL, extracols = NULL)
  
coverpage <- function(title, intro = NULL, about = NULL, source = NULL, relatedlink = NULL, 
                      relatedtext = NULL, dop = NULL, blank = NULL, names = NULL, email = NULL, 
                      phone = NULL, reuse = NULL, govdept = NULL, gridlines = "Yes",
                      extrafields = NULL, extrafieldsb = NULL, additlinks = NULL, addittext = NULL, 
                      colwid_spec = NULL, order = NULL)
  
addnote <- function(notenumber, notetext, applictabtext = NULL, linktext1 = NULL, 
                    linktext2 = NULL)
  
notestab <- function(contentslink = NULL, gridlines = "Yes", colwid_spec = NULL, extracols = NULL)
  
adddefinition <- function(term, definition, linktext1 = NULL, linktext2 = NULL)
  
definitionstab <- function(contentslink = NULL, gridlines = "Yes", colwid_spec = NULL, 
                           extracols = NULL)
  
savingtables <- function(filename, odsfile = "No", deletexlsx = NULL)
```

## Before using accessibletablesR

Prior to using accessitablesR to generate accessible spreadsheets, you must have the data tables of interest stored as dataframes in the global environment of R. accessibletablesR will only work
where there is one table per worksheet required, it is not designed for multiple tables within a
single worksheet. The columns in the R dataframes must be named as they are wanted in the final
spreadsheet. The R dataframes should be sorted as they are wanted in the final spreadsheet.  
<br>
For this vignette, a dummy dataframe (dummydf) will be used. This dataframe has been generated from
mtcars and will be installed when accessibletablesR is installed.  
<br>
This package will save objects to the global environment as certain objects are required by
different functions within the package. This approach is not ideal and future development of the
package will look to resolve the issue. However, for the time being it is advised to ensure that
prior to using accessibletablesR that none of the following objects are in the global environment:  
<br>

  * zzz_wb_zzz
  * zzz_table_data2_zzz
  * zzz_tabcontents_zzz
  * zzz_notesdf_zzz
  * zzz_definitions_zzz
  * zzz_covernumrow_zzz
  * zzz_autonotes2_zzz
  * zzz_fontsz_zzz
  * zzz_fontszst_zzz
  * zzz_fontszt_zzz
  * Objects with a prefix zzz_ and a suffix _startrow_zzz
  * Objects with a prefix zzz_ and a suffix _tablestart_zzz  

<br>
The default setting of accessibletablesR is to error if any of the above objects are present in
the global environment when trying to run the package. It is possible to change this setting
(see the section on creating a workbook).

## Creating a workbook

The first function to run is named "workbook". If this function is not run first, then the other
functions will not run, unless a workbook is manually created within R.  
<br>
The code below demonstrates how to create a workbook with a cover page, contents page, notes page
and definitions page. The automatic annotation of tables with information on associated notes has
been selected. A title and a creator have been set for within the "Info" section of a spreadsheet.

```{r, eval = FALSE}
accessibletablesR::workbook(
  covertab = "Yes", contentstab = "Yes", notestab = "Yes", 
  definitionstab = "Yes", autonotes = "Yes", 
  title = "Fuel consumption and aspects of car design and performance for various cars",
  creator = "An organisation")
```

All parameters within the "workbook" function are optional. The default setting for most of the
parameters is NULL (e.g., no cover page will be created). The "workbook" function will create a
workbook named zzz_wb_zzz in the R global environment. The parameters "subject" and "category" are 
not used in the example code above. These are related to fields within the "Info" section of a spreadsheet. It is possible for the font, font colour and font sizes (different sizes can be 
specified for subtitles, titles and the rest of the spreadsheet) to be amended. The default colour 
is black, the default font is Arial and the default font sizes are 16 for titles, 14 for subtitles 
and 12 for the rest of the spreadsheet.  
<br>
It is advised to keep the parameter autodel set to NULL. However, if it is desired to be able to run
the package without worrying about existing objects in the global environment then set autodel to 
"Yes". This may be desirable when initially trying out the accessibletablesR package.

## Inserting data tables as worksheets

After the workbook has been created, you can start inserting data tables (stored within the R global
environment as dataframes). Each data table will form a separate worksheet in the final spreadsheet.  
<br>
The code below is creating three worksheets. For the purposes of this example, each spreadsheet will
have the same data. In reality, different dataframes will likely be used for different worksheets.  
<br>
Links to the contents, notes and definitions pages can be generated by setting one of the extralines
parameters to "Link to contents" or "Link to notes" or "Link to definitions". Although there are six
extralines parameters, each one can be a vector and there is no limit to how many rows can appear
above the main data (except the limit of rows that Excel can cope with).  
<br>
The example code will create three worksheets which are identical with the exception of column
widths. Table 1 has its column widths determined automatically by the openxlsx package. The column
widths for Table 2 are determined automatically within the "creatingtables" function by counting
the number of characters within columns plus the value of width_adj. The column widths in Table 3
are specified in the call of the function (colwid_spec).  
<br>
Specifying the position of columns containing data as numbers (numdatacols) enables 
accessibletablesR to format the columns correctly (e.g., right aligning, thousand commas). To
specify the desired number of decimal places for each number data column, populate numdatacolsdp. 
To format columns containing non-number data, populate othdatacols with the column positions. At
present there is an issue formatting dates, so if any data in the form of dates are present, these
may need to be formatted manually.  

```{r, eval = FALSE}
accessibletablesR::creatingtables(
  title = "Fuel consumption and aspects of car design and performance for various cars A",
  subtitle = "Cars",
  extraline1 = "Link to contents",
  extraline2 = "Link to notes",
  extraline3 = "Link to definitions",
  sheetname = "Table_1", table_data = accessibletablesR::dummydf, headrowsize = 40,
  numdatacols = c(2:8,11:13), numdatacolsdp = c(1,0,1,0,2,3,2,0,0,1),
  othdatacols = c(9,10))

accessibletablesR::creatingtables(
  title = "Fuel consumption and aspects of car design and performance for various cars B",
  subtitle = "Cars",
  extraline1 = "Link to contents",
  extraline2 = "Link to notes",
  extraline3 = "Link to definitions",
  sheetname = "Table_2", table_data = accessibletablesR::dummydf, headrowsize = 40,
  numdatacols = c(2:8,11:13), numdatacolsdp = c(1,0,1,0,1,0,1,0,0,0),
  othdatacols = c(9,10), columnwidths = "characters", width_adj = 1)

accessibletablesR::creatingtables(
  title = "Fuel consumption and aspects of car design and performance for various cars C",
  subtitle = "Cars",
  extraline1 = "Link to contents",
  extraline2 = "Link to notes",
  extraline3 = "Link to definitions",
  sheetname = "Table_3", table_data = accessibletablesR::dummydf, headrowsize = 40,
  numdatacols = c(2:8,11:13), numdatacolsdp = c(1,0,1,0,2,1,2,0,0,3),
  othdatacols = c(9,10), columnwidths = "specified",
  colwid_spec = c(18,18,18,15,17,15,12,17,12,13,23,22,12))
```

headrowsize adjusts the height of the row containing the table column names. If the user wants to 
give a name to a table in the final Excel workbook which is different to the tab name (sheetname) 
then populate tablename. If the gridlines are not desired in the final workbook, set 
gridlines = "No". extralines1-6 can be set to hyperlinks if desired. An example of how to do this 
is: 

```{r, eval = FALSE}
extraline5 = "[BBC](https://www.bbc.co.uk)"
```

## Creating a contents page

If a contents page is desired, the contentstab parameter in the "workbook" function must have been
set to "Yes". Run the "contentstable" function after all of the data tables have been converted to
worksheets (i.e., run through the "creatingtables" function). All the parameters in the 
"contentstable" function are optional. 

```{r, eval = FALSE}
accessibletablesR::contentstable()
```

If no gridlines are wanted in the contents page in the final workbook set gridlines = "No". Column widths are determined automatically but the user can specify the widths by populating colwid_spec. Extra columns can be provided. To do so, set extracols = "Yes" and create a dataframe called extracols_contents in the global environment before running the "contentstable" function. The extracols_contents dataframe must contain the desired extra columns and have the same number of rows 
as the contents table.

## Adding notes

Notes can be provided in a dedicated notes page. Run the "addnote" function for as many notes as are needed. The function will create a row in the notes page for each note. notenumber is the number of 
the note and should be written as "note" followed by a number (e.g., note1). notetext is the description associated with the note.  
<br>
notenumber and notetext are the only compulsory parameters. If an additional column is wanted to specify which table (sheet name) a note applies to then populate applictabtext (e.g., 
applictabtext = c("Table_1", "Table_2")).  
<br>
An optional column can be included that provides a link to a piece of information. To do so, 
populate linktext1 and linktext2. For example, set linktext1 = "BBC" and set linktext2 to the 
relevant URL address ("https\://www.bbc.co.uk").  
<br>
To automatically annotate each table with the notes that are applicable to it, ensure that the
autonotes parameter in the "workbook" function has been set to "Yes" and that the applictabtext
parameter in the "addnote" function is populated.

```{r, eval = FALSE}
accessibletablesR::addnote(notenumber = "note1", notetext = "Whatever you want note1 to be.", 
                           applictabtext = c("Table_1", "Table_2"), 
                           linktext1 = "No additional link", linktext2 = NULL) 
accessibletablesR::addnote(notenumber = "note2", notetext = "Whatever you want note2 to be", 
                           applictabtext = "Table_1", linktext1 = "No additional link", 
                           linktext2 = NULL) 
accessibletablesR::addnote(notenumber = "note3", notetext = "Whatever you want note3 to be", 
                           applictabtext = "Table_2", linktext1 = "No additional link", 
                           linktext2 = NULL) 
accessibletablesR::addnote(notenumber = "note4", notetext = "Google is an internet search engine", 
                           applictabtext = "All", linktext1 = "Google", 
                           linktext2 = "https://www.ons.google.co.uk") 
```

## Creating a notes page

Once all the notes have been added, the notes page can be created by running the "notestab" 
function. Run the function after all the data tables have been processed using the "creatingtables" function and after all the notes have been added using the "addnote" function. Ensure that the
notestab parameter in the "workbook" function has been set to "Yes".

```{r, eval = FALSE}
accessibletablesR::notestab()
```

The parameters are optional. If a link to the contents page is not wanted on the notes page set contentslink = "No". If no gridlines are wanted on the notes page then set gridlines = "No". Column widths are determined automatically but can be altered to specific widths by the user in 
colwid_spec. Extra columns can be provided. To do so, set extracols = "Yes" and create a dataframe called extracols_notes in the global environment before running the "notestab" function. The extracols_notes dataframe must contain the desired extra columns and have the same number of rows as the notes table.  
<br>
Example code of how to add an extra column to the notes page is below.

```{r, eval = FALSE}
extracols_notes <- data.frame() %>%
  dplyr::mutate(col1 = "") %>%
  dplyr::add_row(col1 = "Whatever you want1") %>%
  dplyr::add_row(col1 = "Whatever you want2") %>%
  dplyr::add_row(col1 = "Whatever you want3") %>%
  dplyr::add_row(col1 = "Whatever you want4") %>%
  dplyr::rename("Extra column name" = col1)

accessibletablesR::notestab(extracols = "Yes")
```

## Adding definitions

Definitions of terms can be provided in a dedicated definitions page. Run the "adddefinition"
function for as many definitions as are needed. term is the item that needs defining and definition
is the definition of the item. An optional column can be included that provides a link to a piece of information. To do so, populate linktext1 and linktext2. For example, set linktext1 = "BBC" and set linktext2 to the relevant URL address ("https\://www.bbc.co.uk"). 

```{r, eval = FALSE}
accessibletablesR::adddefinition(
  term = "Usual resident", 
  definition = "A usual resident is anyone who, on Census Day, 21 March 2021 was in the UK ....")
```

## Creating a definitions page

Run the "definitionstab" function if a definitions page is wanted. Run the function after all the definitions have been added using the "adddefinition" function. Ensure that the definitionstab 
parameter in the "workbook" function has been set to "Yes".

```{r, eval = FALSE}
accessibletablesR::definitionstab()
```

The parameters are optional. If a link to the contents page is not wanted on the definitions page 
set contentslink = "No". If no gridlines are wanted on the definitions page then set 
gridlines = "No". Column widths are set automatically but can be altered using colwid_spec. Extra columns can be provided. To do so, set extracols = "Yes" and create a dataframe called extracols_definitions in the global environment before running the "definitionstab" function. The extracols_definitions dataframe must contain the desired extra columns and have the same number of 
rows as the definitions table.

## Creating a cover page

To create a cover page for the spreadsheet, run the "coverpage" function. Ensure that the covertab
parameter in the "workbook" function has been set to "Yes".  
<br>
title is the only compulsory parameter. It is the title to be displayed on the cover sheet. Other optional sections of a cover page that can be populated are "Introductory information" (info), 
"About these data" (about), "Source" (source), "Related publications" (relatedlink, relatedtext), 
"Date of publication" (dop), "Blank cells" (blank), "Contact" (names, email, phone), 
"Additional links" (additlinks) and "Reusing this publication" (reuse, govdept). Extra fields can be added using extrafields. One row is allowed for each extra field. The text to populate the extra 
fields can be provided in extrafieldsb. If no gridlines are wanted on the cover page in the final workbook set gridlines = "No". The column width is automatically set but can be altered by using colwid_spec.  
<br>
The ordering of the fields can be amended by populating order. order can be set to a vector where 
the field names are provided in speech marks.  
<br>
The "Reusing this publication" section has been designed for UK government departments and will not apply for other organisations. If a user is from the Office for National Statistics (ONS) and wants 
a "Reusing this publication" section then set reuse = "Yes" and govdept to "ONS" or "Office for
National Statistics". If a user is from a UK government department but not the Office for National Statistics (ONS) set reuse = "Yes" and govdept = "name of organisation".  
<br>
The example code below is a simple use of the "coverpage" function.

```{r, eval = FALSE}
accessibletablesR::coverpage(
  title = "Fuel consumption and aspects of car design and performance for various cars",
  intro = "Some made up data about cars",
  about = "The output of an example of how to use accessibletablesR",
  source = "R mtcars",
  relatedlink = "https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars)",
  relatedtext = "mtcars: Motor trend car road tests",
  dop = "26 October 2023",
  blank = "There should be no blank cells",
  names = "Your name",
  email = "yourname@emailprovider.com",
  phone = "01111 1111111111111",
  reuse = "Yes", govdept = NULL)
```

The example code below is a more complex use of the "coverpage" function. Three extra fields have 
been added (see the parameters extrafields and extrafieldsb). Additional links have been provided 
(see the parameters additlinks and addittext). The width of the column has been set to a specific
value (see the colwid_spec parameter). The ordering of the fields on the cover page has been
altered (see the parameter order). The parameter source has been set to a hyperlink (the text 
"R mtcars" will appear and when clicked on will take the user to the website specified).

```{r, eval = FALSE}
accessibletablesR::coverpage(
  title = "Fuel consumption and aspects of car design and performance for various cars",
  intro = "Some made up data about cars",
  about = "The output of an example of how to use accessibletablesR",
  source = 
     "[R mtcars](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars)",
  relatedlink = "https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars)",
  relatedtext = "mtcars: Motor trend car road tests",
  dop = "26 October 2023",
  blank = "There should be no blank cells",
  names = "Your name",
  email = "yourname@emailprovider.com",
  phone = "01111 1111111111111",
  reuse = "Yes", govdept = NULL,
  extrafields = c("Extra heading 1", "Extra heading2", "Extra heading3"),
  extrafieldsb = c("Another line", "Another line", "Another line"),
  additlinks = c("https://www.bbc.co.uk", "https://google.co.uk"),
  addittext = c("BBC", "Google"),
  colwid_spec = 85,
  order = c("intro", "about", "source", "relatedlink", "names", "email", "phone",
            "blank", "extrafields", "additlinks", "reuse"))
```

## Saving the final spreadsheet

To save the final spreadsheet, run the "savingtables" function. This function should be run last
after all other functions have been run. A xlsx or xls spreadsheet will be saved. It is advisable
to not save as a xls file. Ideally, an ods file should be produced. The default setting of the
function is to only save the xlsx or xls file. This is only due to the time it may take to save both
a xlsx file and an ods file (at present the only way to produce an ods file is to save a xlsx file,
read it back in to R, convert to ods and save the ods file).  
<br>
filename is the location and name of the final workbook.  
<br>
To save just a xlsx file, see the example code below.

```{r, eval = FALSE}
accessibletablesR::savingtables("D:/mtcarsexample.xlsx")
```

To save and keep only an ods file, see the example code below.

```{r, eval = FALSE}
accessibletablesR::savingtables("D:/mtcarsexample.xlsx", odsfile = "Yes")
```

To save and keep both a xlsx file and an ods file, see the example code below.

```{r, eval = FALSE}
accessibletablesR::savingtables("D:/mtcarsexample.xlsx", odsfile = "Yes", deletexlsx = "No")
```

## Contact

Please submit any suggestions or report bugs: https://github.com/minifiemj/accessibletablesR/issues  
Or email me: minifiemj@gmail.com